---
title: "Storm Data Analysis"
author: "Endri Raco"
date: "4/6/2020"
output: html_document
keep_md: true
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if(!require(tidyverse)) install.packages('tidyverse', 
repos = 'http://cran.us.r-project.org')
if(!require(gridExtra)) install.packages('gridExtra', 
repos = 'http://cran.us.r-project.org')
## specify global chunk options
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
                      out.width = '90%', fig.align = 'center',
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE,
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```
## Synopsis

This paper presents some insights regarding effects of storms and other severe weather events on both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

For this research we will explore U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.The database covers the time period between 1950 and November 2011. 

The research aims to address the following questions:

  - Across the United States, which types of events are most harmful with respect to population health?
  
  - Across the United States, which types of events have the greatest economic consequences?

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Methods and Analysis

### Importing data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. 

You can download the file from the course web site:

[Storm Data [47Mb]](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)


Let's start with data download.  We copy link used in project description and assign it to **fileurl**.

```{r, data_download}
# We copy link used in project description and assign it to fileurl
filename <- 'repdata_data_StormData.csv'
# Controll for already existing files
# If folder doesn't exist proceed with download
if (!file.exists(filename)){
  fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
  download.file(fileURL, filename, method='curl')
}  
#If file exists proceed with unzip
if (!file.exists('repdata_data_StormData.csv')) { 
  unzip(filename) 
}
```

Now we load our downloaded data:

```{r, data_read}
# Read repdata_data_StormData.csv
data <- read.csv("repdata_data_StormData.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
names (data)
```

Let's see how are data look like:

```{r, data_summary}
# A first look to data
head(data)
# Let's check structure of dataframe
glimpse(data)
```


### Data Processing

We will focus to thosse observations related to injuries, fatalities, property damage
and crop damage. Also we will turn variable names to lower case

```{r, damage_data}
# Create df with focus data
# Name of variables in the lower case
df <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
rename_all(tolower)
names(df) 
```

Next step is to understand variable type and check for missing values:

```{r, data_check}
# Let's check for the number of missing values
sum(is.na(df)) 
# We check classes of every column inside data
lapply(df,class)
```
There are no missing values in the variables inside **df**.  We have 3 variables of type **character** (evtype, propdmgexp, cropdmgexp).

For **evtype** we will clean cases from leading and trailing whitespaces and dublications to avoid problems with classification:


```{r, upper_case}
# Clean evtype
df<- df %>% mutate(evtype = gsub(" +"," ",trimws(evtype)))
```

Let's check if there are other symbols other than **K, M, B, H** inside variables **propdmgexp, cropdmgexp**

```{r, check_symbols}
# For variable propdmgexp
unique(df$propdmgexp)
# For variable crodmgexp
unique(df$cropdmgexp)
```

Based in *National Weather Service Storm Data Documentation* we have to recode all these symbols to their appropriate values. The result of recoding we will save in 2 new variables **propdmgcost, cropdmgcost**, to represent damage created in dollars for properties and damage created in dollars for crop.

We will also create a new variable **propcroptotal** to aggregrate total damage.

Let's start process step by step.


```{r, clean_symbols_prop}
# Create 3 new variables and initiate them
df$propdmgcost=0
df$cropdmgcost=0
df$economy_damage_total =0

# Let's fill new variable propdmgcost with data from recoding of propdmgexp
df[df$propdmgexp=="H",]$propdmgcost=df[df$propdmgexp=="H",]$propdmg*100
df[df$propdmgexp=="K",]$propdmgcost=df[df$propdmgexp=="K",]$propdmg*1000
df[df$propdmgexp=="M",]$propdmgcost=df[df$propdmgexp=="M",]$propdmg*1000000
df[df$propdmgexp=="B",]$propdmgcost=df[df$propdmgexp=="B",]$propdmg*1000000000

# Let's fill new variable cropdmgcost with data from recoding of cropdmgexp
df[df$cropdmgexp=="H",]$cropdmgcost=df[df$cropdmgexp=="H",]$cropdmg*100
df[df$cropdmgexp=="K",]$cropdmgcost=df[df$cropdmgexp=="K",]$cropdmg*1000
df[df$cropdmgexp=="M",]$cropdmgcost=df[df$cropdmgexp=="M",]$cropdmg*1000000
df[df$cropdmgexp=="B",]$cropdmgcost=df[df$cropdmgexp=="B",]$cropdmg*1000000000
# Total cost
df$totalcostdmg = df$propdmgcost + df$cropdmgcost
## Aggregate cost of damages
economic_effect <- df %>% 
select(evtype, totalcostdmg) %>%
    group_by(evtype) %>%
    summarise(costdmg = sum(totalcostdmg)) %>% 
    arrange(desc(costdmg))
    head(economic_effect, 10)
```


### Data Analysis

Now let's see if we can answer to question about types of events (as indicated in the \color{red}{\verb|evtype|}Evtype} variable) are most harmful with respect to population health across the United States ?

For this we aggregrate **fatalities**  by **evtype** to calculate sums and see top values

```{r, agg_fatalities}
## Aggregate fatalities
agg_fatalities <- select(df, evtype, fatalities) %>%
    group_by(evtype) %>%
    summarise(sum_fatalities = sum(fatalities)) %>% 
    arrange(desc(sum_fatalities))
    head(agg_fatalities, 10)
```

Let's visualize our numeric results regarding top fatalities:


```{r, plot_fatalities}
## Plot top fatalities
top_fatalities <- ggplot(head(agg_fatalities, 10), aes(x=reorder(evtype, sum_fatalities), y=sum_fatalities)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Fatalities") + xlab("Event") +
    ggtitle("Most harmful weather events by fatality") +
    theme(legend.position="none")
top_fatalities
```

Let's repeat our previous steps for identifying top injuries

```{r, top_injuries}
## Aggregate injuries        
agg_injuries <- select(df, evtype, injuries) %>% 
      group_by(evtype) %>% 
       summarise(sum_injuries = sum(injuries)) %>% 
       arrange(desc(sum_injuries))
       head(agg_injuries, 10)
```

And the plot:

```{r, plot_injuries}
## Plot top injuries
top_injuries <- ggplot(head(agg_injuries, 10), aes(x=reorder(evtype, sum_injuries), y=sum_injuries)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Injuries") + xlab("Event") +
    ggtitle("Most harmful weather events by injuries") +
    theme(legend.position="none")
top_injuries
```


When it comes to economic consequences, We will vizualize top 10 factors judging by damages in dollars.

```{r, cost_dmg}
## Plot top events which damages have highest cost
top_costdmg <- ggplot(head(economic_effect, 10), aes(x=reorder(evtype, costdmg), y=costdmg)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Cost in dollars") + xlab("Event") +
    ggtitle("Top events which damages have highest cost") +
    theme(legend.position="none")
top_costdmg 
```
Now let's replot the above plot using logarithmic scale. We will put both plots in the same grid to compare if the used scale helps in inproving understanding.

```{r, logplot}
## Plot top events which damages have highest cost
top_costdmg <- ggplot(head(economic_effect, 10), aes(x=reorder(evtype, costdmg), y=costdmg)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Cost in dollars") + xlab("Event") +
    ggtitle("Top events which damages have highest cost") +
    theme(legend.position="none")
top_costdmg 
## Plot the same plot using logarithmic scale
top_costdmg_log <- ggplot(head(economic_effect, 10), aes(x=reorder(evtype, costdmg), y=log10(costdmg))) +
   geom_bar(fill="yellow",stat="identity")  + coord_flip() + 
    ylab("Cost in dollars(log scale)") + xlab("Event") +
    ggtitle("Top events which damages have highest cost") +
    theme(legend.position="none")
top_costdmg_log
## Arrange in grid
grid.arrange(top_costdmg, top_costdmg_log, ncol=1, nrow =2)
```

## Conclusions

Analysis shows that Tornados and Excessive Heat are the top weather events when it comes to fatalities.

Tornados are also the top weather event with respect to injuries. Suprising fact is that no matter how fatal can be Excessive Heat, this event is only in the fourth place when it comes to injuries, and is precedded by Thunderstorms and Flood.

Events that mostly affect economics and largest property damages are flood and hurricanes.
