---
title: "Storm Data Analysis"
author: "Endri Raco"
date: "4/6/2020"
output: html_document
keep_md: true
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if(!require(tidyverse)) install.packages('tidyverse', 
repos = 'http://cran.us.r-project.org')
if(!require(lattice)) install.packages('lattice', 
repos = 'http://cran.us.r-project.org')
## specify global chunk options
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
                      out.width = '90%', fig.align = 'center',
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE,
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Methods and Analysis

### Importing data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. 

You can download the file from the course web site:

[Storm Data [47Mb]](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)


Let's start with data download.  We copy link used in project description and assign it to **fileurl**.

```{r, data_download}
# We copy link used in project description and assign it to fileurl
filename <- 'repdata_data_StormData.csv'
# Controll for already existing files
# If folder doesn't exist proceed with download
if (!file.exists(filename)){
  fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
  download.file(fileURL, filename, method='curl')
}  
#If file exists proceed with unzip
if (!file.exists('repdata_data_StormData.csv')) { 
  unzip(filename) 
}
```

Now we load our downloaded data:

```{r, data_read}
# Read repdata_data_StormData.csv
data <- read.csv("repdata_data_StormData.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
names (data)
```

Let's see how are data look like:

```{r, data_summary}
# A first look to data
head(data)
# Let's check structure of dataframe
glimpse(data)
```




### Data Processing

We will focus to thosse observations related to injuries, fatalities, property damage
and crop damage. Also we will turn variable names to lower case

```{r, damage_data}
# Create df with focus data
# Name of variables in the lower case
df <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
rename_all(tolower)
names(df) 
```

Next step is to understand variable type and check for missing values:

```{r, data_check}
# Let's check for the number of missing values
sum(is.na(df)) 
# We check classes of every column inside data
lapply(df,class)
```
There are no missing values in the variables inside **df**.  We have 3 variables of type **character** (evtype, propdmgexp, cropdmgexp).

In order not to create confusion while reporting results we convert all cases inside these variables in upper case. 

For **evtype** we will clean cases from leading and trailing whitespaces and dublications to avoid problems with classification:

```{r, upper_case}
# Convert to upper case all variables of type character
df <- data.frame(lapply(df, function(v) {
  if (is.character(v)) return(toupper(v))
  else return(v)
}))
# Clean evtype
df<- df %>% mutate(evtype = gsub(" +"," ",trimws(evtype)))
```

In *National Weather Service Storm Data Documentation* is explained that  characters **K, M, B** have the respective meaning for thousands (K), millions(M), and billions(B). 

Let's check if there are other symbols other than **K, M, B** inside variables **propdmgexp, cropdmgexp**

```{r, check_symbols}
# For variable propdmgexp
unique(df$propdmgexp)
# For variable crodmgexp
unique(df$cropdmgexp)
```

We have to remove all symbols other than K, M, B from both variables.

```{r, clean_symbols}
df <- df %>% 
filter(cropdmgexp %in% c("K","M","B","") & propdmgexp %in% c("K","M","B",""))
```

Let's check our results:

```{r, clean_res}
unique(df$propdmgexp)
unique(df$cropdmgexp)
```


### Data Analysis

Now let's see if we can answer to question about types of events (as indicated in the \color{red}{\verb|evtype|}Evtype} variable) are most harmful with respect to population health across the United States ?

For this we aggregrate **fatalities**  by **evtype** to calculate sums and see top values

```{r, agg_injuries}
## Aggregate fatalities
agg_fatalities <- select(df, evtype, fatalities) %>%
    group_by(evtype) %>%
    summarise(sum_fatalities = sum(fatalities)) %>% 
    arrange(desc(sum_fatalities))
    head(agg_fatalities, 10)
```

Let's visualize our numeric results regarding top fatalities:


```{r, agg_injuries}
## Plot top fatalities
top_fatalities <- ggplot(head(agg_fatalities, 10), aes(x=reorder(evtype, sum_fatalities), y=sum_fatalities)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Fatalities") + xlab("Event") +
    ggtitle("Most harmful weather events by fatality") +
    theme(legend.position="none")
top_fatalities
```

Let's repeat our previous steps for identifying top injuries

```{r, top_injuries}
## Aggregate injuries        
agg_injuries <- select(df, evtype, injuries) %>% 
      group_by(evtype) %>% 
       summarise(sum_injuries = sum(injuries)) %>% 
       arrange(desc(sum_injuries))
       head(agg_injuries, 10)
```

And the plot:

```{r, plot_injuries}
## Plot top injuries
top_injuries <- ggplot(head(agg_injuries, 10), aes(x=reorder(evtype, sum_injuries), y=sum_injuries)) +
   geom_bar(fill="green",stat="identity")  + coord_flip() + 
    ylab("Injuries") + xlab("Event") +
    ggtitle("Most harmful weather events by injuries") +
    theme(legend.position="none")
top_injuries
```


When it comes to economic consequences, We will vizualize top 10 factors judging by damages in dollars.



## Conclusions

